---
title: "Home Credit Default Risk — Exploratory Data Analysis"
author: "Anderson Silva"
date: "2026-09-20"
format:
  html:
    toc: true
    toc-title: "Contents"
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

# Business Problem

Home Credit needs to distinguish applicants who are more likely to experience payment difficulty from applicants who are likely to repay. Both decision errors matter: approving higher-risk applicants can increase credit losses, while rejecting applicants who could repay can reduce access to credit and potential business value. `TARGET = 1` is the observed payment-difficulty outcome and `TARGET = 0` means no observed payment difficulty.

Later stages of the project will use supervised classification and decision thresholds. This notebook focuses on exploratory data analysis (EDA): understanding the application data, identifying data-quality issues, testing which applicant characteristics are associated with payment difficulty, and converting those findings into clear data-preparation decisions.

This version focuses on `application_train.csv` and `application_test.csv`. The supplementary Home Credit tables are not joined here, so the analysis does not measure full credit-file depth, previous repayment behavior, or true thin-file status.

# Questions

The EDA is organized around six data-integrity questions and five business questions. The final business question summarizes the strongest descriptive associations found in the earlier sections.

1. Is the target distribution sufficiently represented for EDA and later modeling?
2. Is missingness related to the observed payment-difficulty outcome?
3. Are there impossible, undocumented, or implausible values that require a data-preparation decision?
4. Do the two application tables have the expected grain, keys, and duplicate structure?
5. Do relative-time variables follow the direction described in the data dictionary?
6. Are the train and test application tables structurally and distributionally consistent enough for a common modeling workflow?
7. Is payment difficulty associated with application-level proxies for the burden of the proposed loan relative to reported income?
8. How well do the external risk scores distinguish applicants who experience payment difficulty from those who do not?
9. Is greater recent credit-seeking activity associated with higher payment difficulty?
10. Which available indicators of income level/source and current employment context or tenure are associated with payment difficulty?
11. Which applicant characteristics are most strongly associated with payment difficulty?

```{r}
#| label: setup

# Main libraries used in this notebook.
library(dplyr)
library(ggplot2)

# Keep large numbers readable in the output.
options(scipen = 999)

# Load only the files required for this EDA.
train <- read.csv("../data/raw/application_train.csv")
test <- read.csv("../data/raw/application_test.csv")
dictionary <- read.csv("../docs/HomeCredit_columns_description.csv")
```

# Data Description

Both application tables have the same grain: one row represents one current loan application identified by `SK_ID_CURR`. The training table contains `TARGET`; the test table does not. `SK_ID_CURR` is unique within each table, and the ID sets do not overlap. The two tables are therefore separate samples, not tables that should be joined to each other.

```{r}
# Quick structural summary of the two application tables.
data_structure <- data.frame(
  table = c("application_train.csv", "application_test.csv"),
  rows = c(nrow(train), nrow(test)),
  columns = c(ncol(train), ncol(test)),
  unique_SK_ID_CURR = c(length(unique(train$SK_ID_CURR)),
                        length(unique(test$SK_ID_CURR))),
  duplicated_keys = c(sum(duplicated(train$SK_ID_CURR)),
                      sum(duplicated(test$SK_ID_CURR))),
  TARGET_present = c("Yes", "No")
)

knitr::kable(data_structure, caption = "Application-table structure")
```

For the analysis, I grouped the application columns by the type of information they represent.

| Column group | Examples | Role in this EDA |
|---|---|---|
| Identifier and outcome | `SK_ID_CURR`, `TARGET` | Application key and observed payment-difficulty outcome |
| Loan and income amounts | `AMT_INCOME_TOTAL`, `AMT_CREDIT`, `AMT_ANNUITY`, `AMT_GOODS_PRICE` | Proposed-loan burden relative to reported income |
| Applicant and household | `DAYS_BIRTH`, `CNT_CHILDREN`, `CNT_FAM_MEMBERS`, `NAME_FAMILY_STATUS`, `NAME_EDUCATION_TYPE` | Basic applicant profile |
| Income and employment | `NAME_INCOME_TYPE`, `DAYS_EMPLOYED`, `OCCUPATION_TYPE`, `ORGANIZATION_TYPE` | Current income and employment context |
| External risk scores | `EXT_SOURCE_1`, `EXT_SOURCE_2`, `EXT_SOURCE_3` | External normalized risk information |
| Housing and property | `NAME_HOUSING_TYPE`, `FLAG_OWN_REALTY`, building `*_AVG`, `*_MODE`, `*_MEDI` fields | Housing and building information |
| Contact, region, and documents | `FLAG_*`, `REG_*`, `LIVE_*`, `REGION_RATING_*`, `FLAG_DOCUMENT_*` | Contact, location, and application-document information |
| Bureau inquiries | `AMT_REQ_CREDIT_BUREAU_*` | Recent credit-seeking activity |

The data map also identified several discrepancies between the dictionary and the observed values. `DAYS_EMPLOYED = 365243` is not documented, `CODE_GENDER` contains `XNA`, and the test set contains one `REGION_RATING_CLIENT_W_CITY = -1`. Four building `*_MODE` fields are described as normalized information even though their observed values are categorical. The dictionary also lists `TARGET` under `application_{train|test}`, although the test table correctly has no target.

# Data Integrity

## 1. Target Distribution

**Question:** Is the target distribution sufficiently represented for EDA and later modeling?

```{r}
# Count the two TARGET outcomes and calculate their percentage.
target_distribution <- train |>
  count(TARGET, name = "n") |>
  mutate(percent = n / nrow(train) * 100)

knitr::kable(
  target_distribution,
  digits = 2,
  caption = "TARGET distribution in application_train"
)

ggplot(target_distribution,
       aes(x = factor(TARGET), y = percent)) +
  geom_col() +
  labs(
    title = "Distribution of the Observed Payment-Difficulty Outcome",
    x = "TARGET",
    y = "Applicants (%)"
  ) +
  theme_minimal()
```

`TARGET = 1` represents about 8% of the training applications, so the outcome is materially imbalanced. The sample is still large enough for EDA, but later validation should preserve the class proportion. Model evaluation should not rely on accuracy alone; ROC-AUC and threshold-based tradeoffs are more consistent with the Business Problem Statement (BPS).

## 2. Missingness and `TARGET`

**Question:** Is missingness related to the observed payment-difficulty outcome?

```{r}
# Missingness across every training column.
missing_summary <- data.frame(
  variable = names(train),
  missing_n = sapply(train, function(x) sum(is.na(x))),
  missing_pct = sapply(train, function(x) mean(is.na(x)) * 100)
) |>
  filter(missing_n > 0) |>
  arrange(desc(missing_pct))

# Show only the 15 columns with the most missing data.
knitr::kable(
  head(missing_summary, 15),
  digits = 2,
  caption = "Columns with the highest missingness"
)
```

```{r}
# For each column, compare TARGET when the value is missing versus present.
# A minimum of 1,000 rows in each group avoids interpreting very small samples.
missing_target <- bind_rows(lapply(setdiff(names(train), "TARGET"), function(v) {
  missing <- is.na(train[[v]])

  if (sum(missing) == 0 || sum(!missing) == 0) return(NULL)

  data.frame(
    variable = v,
    missing_n = sum(missing),
    present_n = sum(!missing),
    missing_pct = mean(missing) * 100,
    target_rate_missing = mean(train$TARGET[missing]) * 100,
    target_rate_present = mean(train$TARGET[!missing]) * 100,
    difference_pp = (mean(train$TARGET[missing]) -
                     mean(train$TARGET[!missing])) * 100
  )
})) |>
  filter(missing_n >= 1000, present_n >= 1000) |>
  arrange(desc(abs(difference_pp)))

knitr::kable(
  head(missing_target, 15),
  digits = 2,
  caption = "Largest stable TARGET-rate differences for missing versus present values"
)
```

Missingness is related to `TARGET` for some variables. The six bureau-inquiry fields, for example, are missing for the same 13.5% of applications, and those applicants have a higher payment-difficulty rate than applicants with inquiry information. Complete-case deletion would therefore remove potentially informative cases. The preparation workflow preserves missingness indicators for the external scores and bureau-inquiry fields, while other missing-data treatment remains a modeling decision.

## 3. Invalid and Implausible Values

**Question:** Are there impossible, undocumented, or implausible values that require a data-preparation decision?

```{r}
# Direct conflicts or undocumented values identified against the dictionary.
dictionary_issues <- data.frame(
  table = c(
    "train", "test", "test", "train",
    "train/test", "train/test", "train/test", "train/test"
  ),
  variable = c(
    "DAYS_EMPLOYED", "DAYS_EMPLOYED",
    "REGION_RATING_CLIENT_W_CITY", "CODE_GENDER",
    "FONDKAPREMONT_MODE", "HOUSETYPE_MODE",
    "WALLSMATERIAL_MODE", "EMERGENCYSTATE_MODE"
  ),
  finding = c(
    paste(sum(train$DAYS_EMPLOYED == 365243, na.rm = TRUE), "rows = 365243"),
    paste(sum(test$DAYS_EMPLOYED == 365243, na.rm = TRUE), "rows = 365243"),
    paste(sum(test$REGION_RATING_CLIENT_W_CITY == -1, na.rm = TRUE), "row = -1"),
    paste(sum(train$CODE_GENDER == "XNA", na.rm = TRUE), "rows = XNA"),
    "Categorical values although described as normalized information",
    "Categorical values although described as normalized information",
    "Categorical values although described as normalized information",
    "Categorical values although described as normalized information"
  )
)

knitr::kable(dictionary_issues, caption = "Dictionary conflicts and undocumented values")
```

```{r}
# Screening flags for unusually large values.
# These are review flags, not automatic rules for deleting rows.
annuity_income <- train$AMT_ANNUITY / train$AMT_INCOME_TOTAL
credit_income <- train$AMT_CREDIT / train$AMT_INCOME_TOTAL

extreme_values <- data.frame(
  check = c(
    "Reported income",
    "Annuity / income",
    "Credit / income",
    "Quarterly bureau inquiries",
    "Annual bureau inquiries",
    "Children count",
    "Family-member count"
  ),
  observed_max = c(
    max(train$AMT_INCOME_TOTAL, na.rm = TRUE),
    max(annuity_income, na.rm = TRUE),
    max(credit_income, na.rm = TRUE),
    max(train$AMT_REQ_CREDIT_BUREAU_QRT, na.rm = TRUE),
    max(train$AMT_REQ_CREDIT_BUREAU_YEAR, na.rm = TRUE),
    max(train$CNT_CHILDREN, na.rm = TRUE),
    max(train$CNT_FAM_MEMBERS, na.rm = TRUE)
  ),
  rows_flagged = c(
    sum(train$AMT_INCOME_TOTAL >= 10000000, na.rm = TRUE),
    sum(annuity_income > 1, na.rm = TRUE),
    sum(credit_income > 20, na.rm = TRUE),
    sum(train$AMT_REQ_CREDIT_BUREAU_QRT > 20, na.rm = TRUE),
    sum(train$AMT_REQ_CREDIT_BUREAU_YEAR > 20, na.rm = TRUE),
    sum(train$CNT_CHILDREN > 10, na.rm = TRUE),
    sum(train$CNT_FAM_MEMBERS > 10, na.rm = TRUE)
  )
)

knitr::kable(extreme_values, digits = 3, caption = "Values flagged for source verification")
```

`DAYS_EMPLOYED = 365243` cannot represent employment tenure literally, so it should be converted to missing before tenure is calculated. The test-set region rating of `-1` and `CODE_GENDER = XNA` should also be treated as invalid or unknown rather than mapped to valid categories. The four building `*_MODE` fields should remain categorical. Extreme income values, burden ratios, inquiry counts, and household counts are retained unless there is evidence that they are invalid; extremeness alone is not a reason to delete a row.

## 4. Keys, Duplicates, and Grain

**Question:** Do the two application tables have the expected grain, keys, and duplicate structure?

```{r}
# Confirm one row per SK_ID_CURR and check whether the samples overlap.
key_check <- data.frame(
  table = c("application_train", "application_test"),
  rows = c(nrow(train), nrow(test)),
  unique_keys = c(n_distinct(train$SK_ID_CURR),
                  n_distinct(test$SK_ID_CURR)),
  duplicate_keys = c(sum(duplicated(train$SK_ID_CURR)),
                     sum(duplicated(test$SK_ID_CURR)))
)

shared_ids <- length(intersect(train$SK_ID_CURR, test$SK_ID_CURR))

knitr::kable(key_check, caption = "Primary-key check")
cat("Shared SK_ID_CURR values between train and test:", shared_ids)
```

Both tables have one row per application, and `SK_ID_CURR` is unique in each table. The ID sets do not overlap, so there is no train-to-test join relationship. The tables remain separate samples, and target-based EDA uses `application_train` because it is the only table with `TARGET`.

## 5. Relative-Time Direction

**Question:** Do relative-time variables follow the direction described in the data dictionary?

```{r}
# Check every DAYS_* field in the training table.
day_cols <- grep("^DAYS_", names(train), value = TRUE)

time_check <- data.frame(
  variable = day_cols,
  positive_values = sapply(train[day_cols], function(x) sum(x > 0, na.rm = TRUE)),
  min = sapply(train[day_cols], min, na.rm = TRUE),
  max = sapply(train[day_cols], max, na.rm = TRUE)
)

knitr::kable(time_check, caption = "Direction check for DAYS_* variables")
```

The relative-day fields follow the expected non-positive direction except for `DAYS_EMPLOYED = 365243`. For interpretation and feature engineering, valid negative day counts can be converted to positive years. The employment sentinel must first be set to missing rather than interpreted as roughly 1,000 years of employment.

## 6. Train/Test Consistency

**Question:** Are the train and test application tables structurally and distributionally consistent enough for a common modeling workflow?

```{r}
# Compare columns. TARGET is expected to exist only in train.
train_only_cols <- setdiff(names(train), names(test))
test_only_cols <- setdiff(names(test), names(train))

cat("Train-only columns:", paste(train_only_cols, collapse = ", "), "\n")
cat("Test-only columns:", paste(test_only_cols, collapse = ", "), "\n")
```

```{r}
# Check categorical levels that occur in only one sample.
common_character <- intersect(
  names(train)[sapply(train, is.character)],
  names(test)[sapply(test, is.character)]
)

level_differences <- bind_rows(lapply(common_character, function(v) {
  train_levels <- unique(na.omit(train[[v]]))
  test_levels <- unique(na.omit(test[[v]]))

  only_train <- setdiff(train_levels, test_levels)
  only_test <- setdiff(test_levels, train_levels)

  if (length(only_train) == 0 && length(only_test) == 0) return(NULL)

  data.frame(
    variable = v,
    train_only_levels = paste(only_train, collapse = ", "),
    test_only_levels = paste(only_test, collapse = ", ")
  )
}))

knitr::kable(level_differences, caption = "Categorical levels not shared by train and test")
```

```{r}
# Compare a small set of important numeric variables used in the EDA.
consistency_cols <- c(
  "AMT_INCOME_TOTAL", "AMT_CREDIT", "AMT_ANNUITY",
  "EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3",
  "DAYS_BIRTH", "AMT_REQ_CREDIT_BUREAU_YEAR"
)

numeric_consistency <- bind_rows(lapply(consistency_cols, function(v) {
  data.frame(
    variable = v,
    train_median = median(train[[v]], na.rm = TRUE),
    test_median = median(test[[v]], na.rm = TRUE),
    train_missing_pct = mean(is.na(train[[v]])) * 100,
    test_missing_pct = mean(is.na(test[[v]])) * 100
  )
}))

knitr::kable(
  numeric_consistency,
  digits = 3,
  caption = "Train/test comparison for selected numeric variables"
)
```

The two tables have the same application schema except for the expected train-only `TARGET`. A few categorical levels appear only in train, and some variables differ in missingness or central values. A common workflow is still appropriate, but model-specific preprocessing must safely handle absent or unseen categorical levels. Matching column names alone does not imply identical distributions.

# Results

The following sections address the five business questions. A small working copy is used for derived variables so the original application tables remain unchanged.

```{r}
# Create interpretable age and employment-tenure measures.
# The DAYS_EMPLOYED sentinel is converted to NA only in this EDA working copy.
eda <- train |>
  mutate(
    age_years = -DAYS_BIRTH / 365.25,
    DAYS_EMPLOYED_CLEAN = ifelse(DAYS_EMPLOYED == 365243, NA, DAYS_EMPLOYED),
    tenure_years = -DAYS_EMPLOYED_CLEAN / 365.25
  )
```

## 7. Proposed Loan Burden

**Question:** Is payment difficulty associated with application-level proxies for the burden of the proposed loan relative to reported income?

```{r}
# Build three simple burden ratios.
burden <- eda |>
  transmute(
    TARGET,
    annuity_income = AMT_ANNUITY / AMT_INCOME_TOTAL,
    credit_income = AMT_CREDIT / AMT_INCOME_TOTAL,
    goods_income = AMT_GOODS_PRICE / AMT_INCOME_TOTAL
  )

# Create quintiles and calculate the observed TARGET rate for each ratio.
annuity_q <- burden |>
  filter(!is.na(annuity_income), is.finite(annuity_income)) |>
  mutate(band = ntile(annuity_income, 5)) |>
  group_by(band) |>
  summarise(
    n = n(),
    payment_difficulty_rate = mean(TARGET) * 100,
    min_value = min(annuity_income),
    max_value = max(annuity_income),
    .groups = "drop"
  ) |>
  mutate(measure = "Annuity / income", .before = 1)

credit_q <- burden |>
  filter(!is.na(credit_income), is.finite(credit_income)) |>
  mutate(band = ntile(credit_income, 5)) |>
  group_by(band) |>
  summarise(
    n = n(),
    payment_difficulty_rate = mean(TARGET) * 100,
    min_value = min(credit_income),
    max_value = max(credit_income),
    .groups = "drop"
  ) |>
  mutate(measure = "Credit / income", .before = 1)

goods_q <- burden |>
  filter(!is.na(goods_income), is.finite(goods_income)) |>
  mutate(band = ntile(goods_income, 5)) |>
  group_by(band) |>
  summarise(
    n = n(),
    payment_difficulty_rate = mean(TARGET) * 100,
    min_value = min(goods_income),
    max_value = max(goods_income),
    .groups = "drop"
  ) |>
  mutate(measure = "Goods price / income", .before = 1)

burden_summary <- bind_rows(annuity_q, credit_q, goods_q)

knitr::kable(
  burden_summary,
  digits = 3,
  caption = "Payment-difficulty rate by proposed-loan burden quintile"
)
```

```{r}
ggplot(burden_summary,
       aes(x = factor(band), y = payment_difficulty_rate,
           group = measure, linetype = measure)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Payment-Difficulty Rate by Proposed-Loan Burden",
    x = "Quintile (1 = lowest ratio)",
    y = "Payment-difficulty rate (%)",
    linetype = NULL
  ) +
  theme_minimal()
```

The three ratios do not show a simple monotonic relationship with payment difficulty. The annuity-to-income rate rises through the fourth quintile, while the highest credit-to-income and goods-price-to-income groups have lower observed rates than some middle groups. This may reflect underwriting or loan-size selection rather than a direct affordability effect. The ratios remain useful candidate features, but they should not be treated as approval rules or as substitutes for total debt-to-income or residual cash flow, which cannot be calculated from the current application tables.

## 8. External Risk Scores

**Question:** How well do the external risk scores distinguish applicants who experience payment difficulty from those who do not?

```{r}
# Small helper to calculate a rank-based AUC for one score.
# Because lower EXT_SOURCE values are associated with TARGET = 1,
# the direction-adjusted AUC is reported below.
rank_auc <- function(target, score) {
  keep <- !is.na(target) & !is.na(score)
  target <- target[keep]
  score <- score[keep]

  n1 <- sum(target == 1)
  n0 <- sum(target == 0)
  score_rank <- rank(score, ties.method = "average")

  (sum(score_rank[target == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}

ext_vars <- c("EXT_SOURCE_1", "EXT_SOURCE_2", "EXT_SOURCE_3")

ext_summary <- bind_rows(lapply(ext_vars, function(v) {
  raw_auc <- rank_auc(eda$TARGET, eda[[v]])

  data.frame(
    variable = v,
    nonmissing_n = sum(!is.na(eda[[v]])),
    missing_pct = mean(is.na(eda[[v]])) * 100,
    median_TARGET_0 = median(eda[[v]][eda$TARGET == 0], na.rm = TRUE),
    median_TARGET_1 = median(eda[[v]][eda$TARGET == 1], na.rm = TRUE),
    direction_adjusted_AUC = max(raw_auc, 1 - raw_auc)
  )
}))

knitr::kable(ext_summary, digits = 3, caption = "External-score summary")
```

```{r}
# Create score deciles separately so that D1 always means the lowest score.
ext_deciles <- bind_rows(lapply(ext_vars, function(v) {
  eda |>
    filter(!is.na(.data[[v]])) |>
    transmute(
      TARGET,
      variable = v,
      score = .data[[v]],
      decile = ntile(.data[[v]], 10)
    ) |>
    group_by(variable, decile) |>
    summarise(
      n = n(),
      payment_difficulty_rate = mean(TARGET) * 100,
      .groups = "drop"
    )
}))

knitr::kable(ext_deciles, digits = 2, caption = "Payment-difficulty rate by external-score decile")

ggplot(ext_deciles,
       aes(x = factor(decile), y = payment_difficulty_rate,
           group = variable, linetype = variable)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Payment-Difficulty Rate by External-Score Decile",
    x = "Decile (1 = lowest score)",
    y = "Payment-difficulty rate (%)",
    linetype = NULL
  ) +
  theme_minimal()
```

All three external scores have an inverse relationship with payment difficulty: lower scores are associated with higher observed rates. In this sample, `EXT_SOURCE_3` shows the strongest standalone separation, followed by `EXT_SOURCE_1` and `EXT_SOURCE_2`. The dictionary does not identify the score providers or construction methods, so the scores should be treated as external risk signals rather than interpreted as FICO-equivalent measures. All three remain priority candidate predictors, with their missingness preserved.

## 9. Recent Credit Seeking

**Question:** Is greater recent credit-seeking activity associated with higher payment difficulty?

```{r}
# Group inquiry counts into simple bands for each lookback window.
inquiry_vars <- c(
  HOUR = "AMT_REQ_CREDIT_BUREAU_HOUR",
  DAY = "AMT_REQ_CREDIT_BUREAU_DAY",
  WEEK = "AMT_REQ_CREDIT_BUREAU_WEEK",
  MON = "AMT_REQ_CREDIT_BUREAU_MON",
  QRT = "AMT_REQ_CREDIT_BUREAU_QRT",
  YEAR = "AMT_REQ_CREDIT_BUREAU_YEAR"
)

inquiry_summary <- bind_rows(lapply(names(inquiry_vars), function(window) {
  v <- inquiry_vars[[window]]

  eda |>
    transmute(TARGET, value = .data[[v]]) |>
    mutate(
      band = case_when(
        is.na(value) ~ "Missing",
        value == 0 ~ "0",
        value == 1 ~ "1",
        value == 2 ~ "2",
        value >= 3 ~ "3+"
      ),
      band = factor(band, levels = c("0", "1", "2", "3+", "Missing"))
    ) |>
    group_by(band) |>
    summarise(
      n = n(),
      applicant_pct = n() / nrow(eda) * 100,
      payment_difficulty_rate = mean(TARGET) * 100,
      .groups = "drop"
    ) |>
    mutate(window = window, .before = 1)
}))

knitr::kable(inquiry_summary, digits = 2, caption = "Payment-difficulty rate by bureau-inquiry band")
```

```{r}
# Plot the three longer lookback windows because the shorter windows are sparse.
ggplot(
  inquiry_summary |>
    filter(window %in% c("MON", "QRT", "YEAR"), band != "Missing"),
  aes(x = band, y = payment_difficulty_rate,
      group = window, linetype = window)
) +
  geom_line() +
  geom_point() +
  labs(
    title = "Payment-Difficulty Rate by Bureau-Inquiry Count",
    x = "Inquiry-count band",
    y = "Payment-difficulty rate (%)",
    linetype = "Lookback"
  ) +
  theme_minimal()
```

The annual inquiry window shows the clearest increase in payment difficulty as inquiry counts rise. The monthly and quarterly patterns are less consistent. Repeated credit seeking may indicate higher credit demand or financial stress, but the windows do not behave the same way. The inquiry measures should therefore remain separate rather than be summed, and their shared missingness should be preserved because missing inquiry data is itself associated with a higher payment-difficulty rate.

## 10. Income and Employment Context

**Question:** Which available indicators of income level/source and current employment context or tenure are associated with payment difficulty?

```{r}
# Reported income quintiles.
income_summary <- eda |>
  filter(!is.na(AMT_INCOME_TOTAL)) |>
  mutate(income_band = ntile(AMT_INCOME_TOTAL, 5)) |>
  group_by(income_band) |>
  summarise(
    n = n(),
    min_income = min(AMT_INCOME_TOTAL),
    max_income = max(AMT_INCOME_TOTAL),
    payment_difficulty_rate = mean(TARGET) * 100,
    .groups = "drop"
  )

knitr::kable(income_summary, digits = 2, caption = "Payment-difficulty rate by income quintile")
```

```{r}
# Income-source categories. Small groups are combined for a stable summary.
income_counts <- table(ifelse(is.na(eda$NAME_INCOME_TYPE), "Missing", eda$NAME_INCOME_TYPE))
small_income <- names(income_counts[income_counts < 1000])

income_type_summary <- eda |>
  mutate(
    income_type_group = ifelse(
      is.na(NAME_INCOME_TYPE), "Missing",
      ifelse(NAME_INCOME_TYPE %in% small_income, "Other (<1000)", NAME_INCOME_TYPE)
    )
  ) |>
  group_by(income_type_group) |>
  summarise(
    n = n(),
    payment_difficulty_rate = mean(TARGET) * 100,
    .groups = "drop"
  ) |>
  arrange(desc(n))

knitr::kable(income_type_summary, digits = 2, caption = "Payment-difficulty rate by income source")
```

```{r}
# Employment tenure. The sentinel is kept as a separate group.
tenure_summary <- eda |>
  mutate(
    tenure_band = case_when(
      DAYS_EMPLOYED == 365243 ~ "Sentinel / unknown",
      tenure_years < 1 ~ "<1 year",
      tenure_years < 3 ~ "1-<3 years",
      tenure_years < 5 ~ "3-<5 years",
      tenure_years < 10 ~ "5-<10 years",
      tenure_years >= 10 ~ "10+ years"
    ),
    tenure_band = factor(
      tenure_band,
      levels = c("<1 year", "1-<3 years", "3-<5 years",
                 "5-<10 years", "10+ years", "Sentinel / unknown")
    )
  ) |>
  group_by(tenure_band) |>
  summarise(
    n = n(),
    payment_difficulty_rate = mean(TARGET) * 100,
    .groups = "drop"
  )

knitr::kable(tenure_summary, digits = 2, caption = "Payment-difficulty rate by employment-tenure band")
```

```{r}
# Occupation and organization type have many levels.
# Very small groups are combined so that the rate comparison is not driven by tiny samples.
occupation_counts <- table(ifelse(is.na(eda$OCCUPATION_TYPE), "Missing", eda$OCCUPATION_TYPE))
small_occupation <- names(occupation_counts[occupation_counts < 1000])

organization_counts <- table(ifelse(is.na(eda$ORGANIZATION_TYPE), "Missing", eda$ORGANIZATION_TYPE))
small_organization <- names(organization_counts[organization_counts < 3000])

employment_groups <- eda |>
  mutate(
    occupation_group = ifelse(
      is.na(OCCUPATION_TYPE), "Missing",
      ifelse(OCCUPATION_TYPE %in% small_occupation,
             "Other (<1000)", OCCUPATION_TYPE)
    ),
    organization_group = ifelse(
      is.na(ORGANIZATION_TYPE), "Missing",
      ifelse(ORGANIZATION_TYPE %in% small_organization,
             "Other (<3000)", ORGANIZATION_TYPE)
    )
  )

occupation_summary <- employment_groups |>
  group_by(occupation_group) |>
  summarise(
    n = n(),
    payment_difficulty_rate = mean(TARGET) * 100,
    .groups = "drop"
  ) |>
  arrange(desc(payment_difficulty_rate))

organization_summary <- employment_groups |>
  group_by(organization_group) |>
  summarise(
    n = n(),
    payment_difficulty_rate = mean(TARGET) * 100,
    .groups = "drop"
  ) |>
  arrange(desc(payment_difficulty_rate))

knitr::kable(occupation_summary, digits = 2, caption = "Payment-difficulty rate by occupation")
knitr::kable(organization_summary, digits = 2, caption = "Payment-difficulty rate by organization type")
```

```{r}
ggplot(tenure_summary |> filter(tenure_band != "Sentinel / unknown"),
       aes(x = tenure_band, y = payment_difficulty_rate, group = 1)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Payment-Difficulty Rate by Employment Tenure",
    x = "Current employment tenure",
    y = "Payment-difficulty rate (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

The highest income quintile has a lower payment-difficulty rate, and longer valid employment tenure shows a clearer decline in the observed rate. Income source, occupation, and organization type also differ across groups. These variables may capture repayment capacity and employment stability, but they are only a snapshot at the time of application. They remain candidate predictors, with the employment sentinel cleaned before tenure is used numerically. Rare-category grouping and categorical encoding are better handled during model preprocessing, where their effect can be evaluated directly. These descriptive differences should not be interpreted as causal effects.

## 11. Strongest Applicant Associations

**Question:** Which applicant characteristics are most strongly associated with payment difficulty?

This final question combines the earlier results using a simple descriptive measure: the difference between the highest and lowest payment-difficulty rate across sufficiently sized groups for each characteristic. The comparison is univariate. It is not a causal estimate and should not be interpreted as independent model feature importance.

```{r}
# Create a reusable long table with selected applicant characteristics.
# Each characteristic uses the same bands already used in the previous sections.
profile_long <- bind_rows(
  eda |>
    filter(!is.na(EXT_SOURCE_1)) |>
    transmute(TARGET, characteristic = "EXT_SOURCE_1",
              group = as.character(ntile(EXT_SOURCE_1, 10))),
  eda |>
    filter(!is.na(EXT_SOURCE_2)) |>
    transmute(TARGET, characteristic = "EXT_SOURCE_2",
              group = as.character(ntile(EXT_SOURCE_2, 10))),
  eda |>
    filter(!is.na(EXT_SOURCE_3)) |>
    transmute(TARGET, characteristic = "EXT_SOURCE_3",
              group = as.character(ntile(EXT_SOURCE_3, 10))),
  employment_groups |>
    transmute(TARGET, characteristic = "Occupation type",
              group = occupation_group),
  employment_groups |>
    transmute(TARGET, characteristic = "Organization type",
              group = organization_group),
  eda |>
    mutate(
      age_band = cut(
        age_years,
        breaks = c(-Inf, 25, 35, 45, 55, 65, Inf),
        labels = c("<25", "25-34", "35-44", "45-54", "55-64", "65+")
      )
    ) |>
    transmute(TARGET, characteristic = "Age", group = as.character(age_band)),
  eda |>
    mutate(
      tenure_group = case_when(
        DAYS_EMPLOYED == 365243 ~ NA_character_,
        tenure_years < 1 ~ "<1 year",
        tenure_years < 3 ~ "1-<3 years",
        tenure_years < 5 ~ "3-<5 years",
        tenure_years < 10 ~ "5-<10 years",
        tenure_years >= 10 ~ "10+ years"
      )
    ) |>
    transmute(TARGET, characteristic = "Employment tenure", group = tenure_group),
  eda |>
    transmute(TARGET, characteristic = "Education type",
              group = as.character(NAME_EDUCATION_TYPE)),
  eda |>
    transmute(TARGET, characteristic = "Housing type",
              group = as.character(NAME_HOUSING_TYPE)),
  eda |>
    mutate(income_band = ntile(AMT_INCOME_TOTAL, 5)) |>
    transmute(TARGET, characteristic = "Income amount",
              group = as.character(income_band)),
  eda |>
    transmute(TARGET, characteristic = "Income type",
              group = as.character(NAME_INCOME_TYPE)),
  eda |>
    mutate(
      children_group = case_when(
        CNT_CHILDREN == 0 ~ "0",
        CNT_CHILDREN == 1 ~ "1",
        CNT_CHILDREN == 2 ~ "2",
        CNT_CHILDREN >= 3 ~ "3+"
      )
    ) |>
    transmute(TARGET, characteristic = "Children count", group = children_group),
  eda |>
    mutate(
      family_group = case_when(
        CNT_FAM_MEMBERS <= 1 ~ "1",
        CNT_FAM_MEMBERS == 2 ~ "2",
        CNT_FAM_MEMBERS == 3 ~ "3",
        CNT_FAM_MEMBERS == 4 ~ "4",
        CNT_FAM_MEMBERS >= 5 ~ "5+"
      )
    ) |>
    transmute(TARGET, characteristic = "Family members", group = family_group),
  eda |>
    transmute(TARGET, characteristic = "Owns car", group = as.character(FLAG_OWN_CAR)),
  eda |>
    transmute(TARGET, characteristic = "Owns realty", group = as.character(FLAG_OWN_REALTY)),
  eda |>
    filter(!is.na(AMT_REQ_CREDIT_BUREAU_YEAR)) |>
    mutate(
      inquiry_group = case_when(
        AMT_REQ_CREDIT_BUREAU_YEAR == 0 ~ "0",
        AMT_REQ_CREDIT_BUREAU_YEAR == 1 ~ "1",
        AMT_REQ_CREDIT_BUREAU_YEAR == 2 ~ "2",
        AMT_REQ_CREDIT_BUREAU_YEAR >= 3 ~ "3+"
      )
    ) |>
    transmute(TARGET, characteristic = "Annual bureau inquiries", group = inquiry_group),
  burden |>
    filter(!is.na(annuity_income), is.finite(annuity_income)) |>
    mutate(group = as.character(ntile(annuity_income, 5))) |>
    transmute(TARGET, characteristic = "Annuity / income", group),
  burden |>
    filter(!is.na(credit_income), is.finite(credit_income)) |>
    mutate(group = as.character(ntile(credit_income, 5))) |>
    transmute(TARGET, characteristic = "Credit / income", group)
)

# Calculate rates by group, keep groups with at least 1,000 applicants,
# and then calculate the spread for each characteristic.
association_rates <- profile_long |>
  filter(!is.na(group)) |>
  group_by(characteristic, group) |>
  summarise(
    n = n(),
    payment_difficulty_rate = mean(TARGET) * 100,
    .groups = "drop"
  ) |>
  filter(n >= 1000)

association_rank <- association_rates |>
  group_by(characteristic) |>
  filter(n() >= 2) |>
  summarise(
    lowest_group = group[which.min(payment_difficulty_rate)],
    lowest_rate = min(payment_difficulty_rate),
    highest_group = group[which.max(payment_difficulty_rate)],
    highest_rate = max(payment_difficulty_rate),
    spread_pp = max(payment_difficulty_rate) - min(payment_difficulty_rate),
    .groups = "drop"
  ) |>
  arrange(desc(spread_pp))

knitr::kable(
  association_rank,
  digits = 2,
  caption = "Largest descriptive payment-difficulty-rate spreads"
)
```

```{r}
# Plot only the ten largest spreads to keep the figure readable.
ggplot(head(association_rank, 10),
       aes(x = reorder(characteristic, spread_pp), y = spread_pp)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Largest Descriptive Associations with Payment Difficulty",
    x = NULL,
    y = "Rate spread (percentage points)"
  ) +
  theme_minimal()
```

The external-source scores have the largest descriptive rate spreads. Applicant-profile and employment variables such as occupation and age also show meaningful differences. This indicates that the application table already contains variables that separate higher- and lower-risk groups before supplementary credit history is added. A large univariate spread does not mean that a variable will remain important after other predictors are included, so the fitted model will be used to evaluate their contribution jointly.

# Conclusion and Data-Preparation Decisions

The EDA shows that the application data contain meaningful risk differentiation. In the descriptive spread comparison, `EXT_SOURCE_3` has about a 16.8 percentage-point difference between its lowest- and highest-risk deciles. `EXT_SOURCE_2` and `EXT_SOURCE_1` follow at about 15.4 and 15.0 points. Occupation type and age also show large descriptive spreads, at about 12.3 and 8.6 percentage points.

Not every relationship is simple. In particular, the proposed-loan burden ratios are not monotonic, so a larger requested-loan-to-income ratio should not be interpreted as directly implying greater observed risk.

The main data-quality findings and their preparation decisions are:

| Data issue | Decision for data preparation |
|---|---|
| `DAYS_EMPLOYED = 365243` | Recode to missing before calculating `tenure_years` |
| `REGION_RATING_CLIENT_W_CITY = -1` in test | Treat as invalid/missing rather than as a valid region rating |
| `CODE_GENDER = XNA` | Treat as invalid/missing rather than mapping it to a valid category |
| Building `*_MODE` fields described as normalized but observed as categories | Keep them categorical |
| Missingness associated with `TARGET` in external-score and bureau-inquiry fields | Avoid complete-case deletion and preserve missingness indicators |
| Extreme but not necessarily impossible values | Retain them unless there is evidence that they are invalid |
| Train-only categorical levels | Handle unseen or absent levels during model preprocessing |
| Nonlinear annuity-to-income pattern | Retain the ratio and a consistent five-group representation |
| Minority `TARGET = 1` outcome | Preserve class proportions during validation and avoid accuracy-only evaluation |

Relative to the BPS, the EDA supports the assumption that the application data can differentiate repayment risk, especially through the external scores and employment/applicant-profile variables. It also changes the preprocessing plan by making nonlinear relationships, informative missingness, categorical consistency, and the employment sentinel explicit parts of the workflow.

The EDA does not test business targets, holdout ROC-AUC improvement, reason codes, approval-threshold scenarios, or model error rates by group. Those questions require a fitted model and later validation. With only the application tables, the project also cannot establish true thin-file status or full affordability because detailed credit history and household cash-flow information are not included.